---
title: "Analyzing the data of rapper: Pitbull's songs from 2004 to 2024"
subtitle: "Analyzing Pitbull’s Music: Trends in Valence, Energy, and Danceability from 2004 to 2024"
author: 
  - Yuanchen Miao
  - Ziqi Zhu
thanks: "Code and data are available at: https://github.com/NevaeH-9/vincentspotify"
date: 21 November 2024
date-format: long
abstract: "This paper explores trends in the valence, energy, and danceability of Pitbull's songs from 2004 to 2024. The analysis examines how the emotional tone (valence) has changed over time, its relationship with energy, and how energy influences danceability. Using data from Spotify, the study identifies key patterns in Pitbull's music, including shifts in mood and energy over time, as well as the optimal energy levels for danceability."
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(dplyr)
library(tibble)
library(knitr)
library(tidyverse)
library(kableExtra)
```

# Introduction {#sec-introduction}
The R programming language [@R] and tidyverse [@tidyverse] were used to clean data, generate tables and create graphs, package knitr [@knitr] and [@kablee] is used for adding caption to table. The data were gathered and downloaded form spotify by using package spotifyr [@spotifyr] and devtools [@devtools].

# Data {#sec-data}
The data used in this study was obtained using the Spotify Web API [@spotify_api] through the spotifyr package in R [@spotifyr], all of the songs from 2004 to 2024 and "album_release_date", "danceability", "valence" and "energy" are chosen to analyze.  This table @tbl-data shows the first several rows of the data we use.
```{r}
#| include: false
#| warning: false
#| message: false
rds <- readRDS("../data/cleaned_pitbull.rds")
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-data

# View the first few rows of the data
rds %>%
  rename(
    "Album Release Date" = album_release_date,
    "Danceability" = danceability,
    "Valence" = valence,
    "Energy" = energy
  ) %>%
  head(3) %>%
  kable(caption = "2004 - 2024 Data of Pitbull's Songs") %>%
  kable_styling(font_size = 10)
```

# Result {#sec-result}

```{r}
#| label: fig-1
#| fig-cap: "Trend in Valence of Pitbull's songs Over Time" 
#| warning: false
#| message: false
#| echo: false
#| fig-pos: "H"
rds <- as_tibble(rds)

rds |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(
    x = album_release_date,
    y = valence,
    group = album_release_date
  )) +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, width = 0.3, height = 0) +
  theme_minimal() +
  labs(
    x = "Album release date",
    y = "valence"
  )
```
Box plot shows the trend in the musical positivity (valence) of Pitbull's songs from 2004 to 2024, positive in general but become extremely neagtive around 2018, showing fluctuations in emotional tone across albums and highlighting shifts toward more upbeat or moody compositions.

```{r}
#| label: fig-2
#| fig-cap: "Trend in Valence of Pitbull's songs Over Time" 
#| warning: false
#| message: false
#| echo: false
#| fig-pos: "H"
rds |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes( x = album_release_date, y = valence)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  theme_minimal() +
  labs(
    x = "Album release date",
    y = "Valence"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

scatter plot of album released and valence, showing a decrease from 2020 to 2018, and an increase after 2018.
```{r}
#| label: fig-3
#| fig-cap: "Relationship between valence and energy" 
#| warning: false
#| message: false
#| echo: false
#| fig-pos: "H"

rds |>
  ggplot(aes( x = valence, y = energy)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  theme_minimal() +
  labs(
    x = "Valence",
    y = "Energy"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```
Scatter plot showing the relationship between energy and valence in Pitbull's songs. The fit line indicates that songs with moderate positivity are associated with higher energy levels, showing a positive correlation when Valence < 0.5, while extremes in valence tend to have lower energy.

```{r}
#| label: fig-4
#| fig-cap: "Relationship between energy and danceability" 
#| warning: false
#| message: false
#| echo: false
#| fig-pos: "H"

rds |>
  ggplot(aes( x = energy, y = danceability)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  theme_minimal() +
  labs(
    x = "Energy",
    y = "Danceability"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```
Scatter plot showing how danceability changes with energy in Pitbull's songs. The fit line highlights that moderately high energy levels maximize danceability(when valence < 0.5), with a decline in danceability for extremely high-energy tracks.

# Discussion {#sec-discussion}
From the graph @fig-1, we can see that In the early years (pre-2010), there is a wide range of valence values, with some songs having very high positivity (close to 1) and others showing lower values, suggesting a mix of upbeat and moody tracks.
As time progresses (between 2010 and 2020), the valence remains fairly consistent, though the range in some albums narrows, indicating a tendency toward maintaining a certain upbeat style.
Around 2020, there seems to be a drop in valence for a few songs, as seen by some outliers with lower valence scores. This could suggest Pitbull released some less positive songs during this period, though most of his music remained relatively upbeat.

When we fit a smoothed line to the graph as shown in @fig-2, the trend is more obvious, suggests that Pitbull's music started off with a relatively high valence (around 0.75) in the early 2000s, and then gradually declined, reaching a low point around 2018, where the valence dipped below 0.5. After that valence increases again, climbing back toward 0.75 until recent years.

Other two graphs that visualize relationships between valence, energy, and danceability of Pitbull's songs @fig-3 and @fig-4 have similar patterns. Both of them shows there is a strong positive correlation between two variables when one varaible is smaller than 0.5. @fig-3 shows that positive songs tend to have higher energy, but beyond a certain point(around 0.75), increasing valence doesn't necessarily result in more energetic. @fig-4 demonstrates that moderately high energy songs(with energy range from 0.5 to 0.8) are the most danceable, but extremely high energy might reduce danceability.

\newpage

# References


